---
title: "Exploring Exponential Distributions"
author: "James Whedbee"
date: "September 25, 2015"
output: pdf_document
---

The purpose of this report is to investigate the exponential distribution. To aid in this investigation, a distribution of means of 40 exponentials was created using 1000 simulations. A lambda value of 0.2 was chosen for all simulations.

```{r,cache=TRUE}
set.seed(416)
expSimul <- NULL
lambda <- 0.2
n <- 40
for(i in 1:1000) expSimul  <-  c(expSimul,mean(rexp(n,lambda)))
```

This report aims to accompish three goals:

###1) Show the sample mean achieved through simulation and compare it to the theoretical mean of the distribution.

The theoretical mean is given by 1/lambda, which is shown below:
```{r}
1/lambda
```
The sample mean is calculated as follows:
```{r}
mean(expSimul)
```
We can see that the values are quite close.

###2) Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.

The theoretical variance is given by 1/(((lambda)^2)*(n)), which is shown below:
```{r}
1/(((lambda)^2)*(n))
```
The sample variance is calculated as follows:
```{r}
var(expSimul)
```
We can see that the values are quite close.

###3) Show that the distribution is approximately normal.

Here we can see the distribution of the mean of the simulations is approximately normal:

```{r, echo=FALSE}
hist(expSimul)
```

In contrast, here is the distribution of exponentials from a single simulation of 1000 values:

```{r,echo=FALSE,cache=TRUE}
oneRun <- rexp(1000,.2)
hist(oneRun)
```
